Mathematical Aspects of the Capital Market

### 2017-08-21 phd/Chapterintroduction.tex
Updated the thesis manuscript.

\chapter{Introduction}\label{ch:intr}
The phenomena of heavy-tailedness is widely observed in many
disciplines of science, for example, phase transition of matter and
black body radiation as studied in physics, neuronal avalanches in
biology, claim sizes of insurance mathematics and stock returns in
finance. The last application is indeed the focus of this thesis. To
discuss the phenomena in precise terms, we introduce the concept of
regular variation.

\section{Regular Variation}\label{sec:intro_regvar}
The concept of regular variation is defined by the following scaling
property: if a function $f$ on $(0, \infty)$ satisfies
\[
\lim_{x \to \infty} {f(c x) \over f(x)} = c^\alpha
\quad
\forall c > 0,
\]
then we say $f$ is regularly varying with index $\alpha$.
$f$ can be written in the form $f(x) = {L(x) \over x^\alpha}$, where
$L(x)$ is a slowly varying function, i.e.
$\lim_{x \to \infty} {L(c x) \over L(x)} = 1$, $\forall c > 0$.
We call a random variable $X$ regularly varying with index
$\alpha \geq 0$ if it satisfies the tail balance condition in the
limit $x \to \infty$.
\[
\P(X > x) \sim p_+ {L(x) \over x^\alpha},\quad
\P(X < -x) \sim p_- {L(x) \over x^\alpha}, \quad
\text{for some } p_+, p_- > 0,\; p_+ + p_- = 1
\]

@@

Clearly, estimating the tail index $\alpha$ of a sequence
$X_1, X_2, \dots$ of regularly varying variables is particularly
important for understanding the behavior of a heavy-tailed series. A
standard method proposed for this purpose is due to Hill
\cite{hill1975simple}:
\begin{equation}
  \label{eq:fhhty}
  \hat \alpha_H = \left[
    {1 \over k} \sum_{i=1}^k \log \left(
      X_{(i)} \over X_{(k+1)}
    \right)
  \right]^{-1},
\end{equation}
where $X_1, X_2, \dots, X_n$ is a sample whose tail index is the
subject of estimation, and $X_{(i)}$ is the $i$-th upper order
statistic of the sample. Several authors have contributed to showing
the weak consistency and asymptotic normality of the estimator
$\hat \alpha_H$, under the assumptions $k \to \infty, k/n \to 0$
as $n \to \infty$. See theorem 6.4.6 of Embrechts et. al.
\cite{embrechts:klueppelberg:mikosch:1997} for details.

@@

Random variables with regularly varying tails have some very nice
features: if $X_1$ and $X_2$ are both regularly varying with indices
$\alpha_1$ and $\alpha_2$, respectively, then $a_1 X_1 + a_2 X_2$,
for $a_1, a_2 > 0$, is regularly varying with index
$\min\{\alpha_1, \alpha_2\}$ (cf. Mikosch and Jessen
\cite{JessenMikosch2006}). Moreover, if $X_1, X_2$ are iid,
$\P(a_1 X_1 + a_2 X_2 > u) \sim \P(a_1 X_1 > u) + \P(a_2 X_2 > u)$.

Now consider $p$ return series
$X_{i,t}$, $i=1,2,\dots, p; t=1,\dots,n$.
Suppose each of these series is a linear combination of $K$ factors
$Y_{j,t}$, $j=1,2,\dots,K$, the $j$-th of which is regularly varying
with index $\alpha_j$. Then by the summation property, each and every
$\{X_{i,t}\}$ is regularly varying with index $\min_{1 \leq j \leq K} \alpha_j$.
In practice a factor $Y_{i,t}$ is estimated as
$\hat Y_{j,t} = \sum_{i=1}^p a_{j, i} X_{i,t}$, where
$(a_{j, 1}, a_{j, 2}, \dots, a_{j, p})^\top$ is the $j$-th eigenvector
of the sample covariance matrix of
$\{X_{i,t}\}, i=1,\dots,p; t=1,\dots,n$, i.e. the eigenvector
associated with the $j$-th largest eigenvalue. 
For this reason, it is important to understand the eigensystem of the
sample covariance matrix of $\{X_{i,t}\}$. This topic is addressed in
chapters \ref{ch:bernoulli} and \ref{ch:extremes} of this thesis.

@@

The notion of a regularly varying strictly stationary process is also of
interest. Originally introduced by Davis and Hsing
\cite{davis:hsing:1995} for univariate processes, the concept was
extended to multivariate processes by Davis and Mikosch
\cite{davis:mikosch:2009b}: An $\reals^d$-valued strictly stationary
process $X_t$ is said to be regularly varying with index $\alpha$, if
for each $h \geq 0$, 
\[
  {
    \P\left[x^{-1}(X_0, X_1, \dots, X_h) \in \cdot \right]
    \over
    \P(|X_0| > x)
  } \overset{v}{\to} \nu_\alpha(\cdot),
  \quad x \to \infty
\]
where $\nu_\alpha$ a non-null Radon measure on
$\reals^{d(h+1)} \setminus \{0\}$ that
is homogeneous of order $-\alpha$, i.e.
$\forall c > 0, \forall S \subset \reals^{d(h+1)}\setminus\{0\}, \nu_\alpha(c S) = c^{-\alpha} \nu_\alpha(S)$. $|\cdot|$ could be any given vector norm.

Basrak and Segers \cite{basrak:segers:2009} proved an equivalence
condition for regular variation of
a strictly stationary process: an $\reals^d$-valued strictly stationary
process $\{X_t\}$ is regularly varying with index $\alpha$ if 
there exists an $\reals^d$-valued sequence $\{\Theta_i\}_{i = 0,1,\ldots}$
and a Pareto($\alpha$) random variable $Z$, i.e.
$\P(Z > x) = x^{-\alpha}, \forall x \geq 1$, independent of
$\{\Theta_i\}$
\[
\P\left[\left.
  x^{-1}(X_0, X_1, \ldots) \in \cdot
  \right| |X_0| > x
  \right]
  \overset{w}{\to}
  \P(Z(\Theta_1, \Theta_2, \ldots) \in \cdot),
  \quad
  x \to \infty.
\]
The sequences $\{\Theta_i\}_{i = 0,1,\ldots}$ and
$\{Z \Theta_i\}_{i = 0,1,\ldots}$ are called the spectral tail process
and the tail process of $\{X_t\}$, respectively.

A recent remarkable result from Mikosch and Wintenberger
\cite{mikosch:wintenberger:2016} regarding
strictly stationary regularly varying processes is quoted below:
\begin{theorem}\label{thm:mikwin:intro}
Let $(Y_t)$ be an $\bbr^d$-valued strictly stationary sequence , $S_n=Y_1+\cdots +Y_n$ and $(a_n)$ be such that $n\,\P(|Y|>a_n)\to 1$.
Also write for $\vep>0$, $\ov Y_t = Y_t \I(|Y_t|\le \vep a_n)$, $\underline Y_t= Y_t-\ov  Y_t $ and
\beao
\ov S_{l,n}= \sum_{t=1}^l \ov Y_t \qquad  \un S_{l,n}= \sum_{t=1}^l \un Y_t\,.
\eeao
Assume the following conditions:
\begin{enumerate}
\item 
$(Y_t)$ is regularly varying with index $\alpha\in (0,2)  \setminus\{1\}$ and spectral tail process $(\Theta_j)$.
\item
A mixing condition holds: there exists an integer \seq\ $m_n\to\infty$ \st\ $k_n= [n/m_n]\to \infty$
and 
\beam\label{eq:chfa:intro}
\E \ex^{it'\underline S_n/a_n} - \Big(\E \ex^{it'\underline S_{m_n,n}/a_n}\Big)^{k_n} \to 0\,,\qquad \nto\,,\qquad t\in\bbr^d\,.
\eeam
\item An anti-clustering condition holds:
\beam\label{eq:acl:intro}
\lim_{l\to\infty} \limsup_{\nto} \P\big(\max_{t=l,\ldots,m_n} |Y_t|>\delta a_n\mid |Y_0|>\delta a_n\big)=0\,,\qquad \delta>0\,
\eeam
for the same sequence $(m_n)$ as in (2).
\item If $\alpha\in (1,2)$, in addition $\E[Y]=0$ and the vanishing small values condition holds: 
\beam\label{eq:vansm:intro}
\lim_{\vep\downarrow 0}\limsup_{\nto} \P\big(a_n^{-1} |\ov S_n-\E [\ov S_n]|>\delta \big)=0\,,\qquad \delta>0\,
\eeam
and $\sum_{i=1}^\infty \E[|\Theta_i|]<\infty$.
\end{enumerate}
Then $a_n^{-1} S_n\std \xi_\alpha$ for an $\alpha$-stable $\bbr^d$-valued
vector $\xi_\alpha$ with log characteristic function
\begin{small}
  \beam\label{eq:chfid:intro}
  \int_0^\infty \E\left[
    \exp\left(i\,y\,t'\sum_{j=0}^\infty \Theta_j\right)-
    \exp\left(i\,y\,t'\sum_{j=1}^\infty \Theta_j\right)
    -i\,y\,t'\I_{(1,2)}(\alpha)
    \right]\,
  d(-y^{\alpha})\,,\qquad t\in\bbr^d\,.
  \eeam
\end{small}
\end{theorem}\noindent 
\begin{remark}\label{rem:case:alpha:1:intro}
If we additionally assume that $Y$ is symmetric, which implies $\E[\ov
  Y]=\mathbf{0}$, then the statement of the theorem also holds for
$\alpha=1$.
\end{remark}

@@

Kesten \cite{kesten:1973} showed that, when $A_t$ is almost
surely non-negative, has no row or column of only zeros, and
$B_t$ is almost surely non-negative and is not equal to the null
vector with probability 1, then the solution $X$ to the equation
$X \overset{d}{=} A X + B$ is regularly varying with a positive index
$\alpha$, assuming the following conditions (M) and (A):
\begin{itemize}
\item Condition (M)
  \begin{enumerate}
  \item The top Lyapunov exponent
    \[
    \gamma = \inf_{n \geq 1} {1 \over n}\E \log \|A_n \cdots A_1\|
    \]
    is negative.
  \item There exists $\alpha > 0$ such that
    \[
    1 = \lambda(\alpha) = \lim_{n \to \infty} {1 \over n} \log \E
    \|A_n \cdots A_1\|^\alpha
    \]
  \item $\E (\|A_1\|^\alpha \log^+\|A_1\|) < \infty$
  \item $\E |B_1|^\alpha < \infty$
  \end{enumerate}
\item Condition (A) : The group generated by
  \[
  \{\log\rho(s): s = A_n \cdots A_1 \text{ for some } n \geq 1\}
  \]
  is dense in $\reals$, where $\rho(s)$ denotes the spectral
  radius of matrix $s$.
\end{itemize}
Upon these conditions, Kesten's theorem gives
\begin{equation}
  \label{eq:jorg}
  u^\alpha \P(u^{-1} X \in \cdot) \overset{v}{\to} \nu_\alpha(\cdot)
\end{equation}
where $\nu_\alpha$ is a non-null Radon measure on
$\reals^d_+ \setminus \{0\}$ with the property
$\forall a > 0, \nu_\alpha(a A) = a^{-\alpha} \nu_\alpha(A)$.

In addition to non-negative matrices, two other classes of random
matrices have been shown to lead to power-law tails via the recurrence
relation \eqref{eq:garchpq_sre}. Alsmeyer and Mentemeier
\cite{alsmeyer:mentemeier:2012} considered invertible matrices whose
distribution has a density. Let $M(d, \reals)$ denote the space of
$d \times d$ matrices with real entries that are invertible with
probability 1. They replaced Kesten's condition
$\E (\|A\|^\alpha \log^+\|A\|) < \infty$ with a stronger
counterpart
$\E [\|A\|^\alpha (\log^+\|A\| + \log \|A^{-1} \|)] < \infty$,
and lifted the condition (A). In addition, they assumed
\begin{enumerate}
  \item For any open set $U \subset \sphere^{d-1}$ and any
    $u \in \sphere^{d-1}$,  $\exists n \geq 1$ such that
    \[
    \P\left({\prod_{i=1} A_i u
        \over
        \left|\prod_{i=1} A_i u\right|}\in U
    \right) > 0.
    \]
  \item There exist $N \geq 1$, $c, \epsilon > 0$ and an invertible
    matrix $\bar A \in M(d, \reals)$ such that for any set
    $C \subset M(d, \reals)$, it holds true
    $\P(A_N \cdots A_1 \in C) \geq c |B_\epsilon(\bar A) \cap C|$,
    where $|\cdot|$ denotes the Lebesgue measure.
\end{enumerate}
These assumptions are termed conditions (id). Furthermore, they assumed
that there was no point in $\reals^d$ such that the recurrence
equation \eqref{eq:garchpq_sre} was stuck at this point with probability 1: 
$\P(A x + B = x) < 1$ for all $x \in \reals^d$. With these
assumptions, they showed
\[
\lim_{u \to \infty} u^\alpha \P(\inn{x, X} > u) = e_\alpha(x)
\]
where $x \in \sphere^{d - 1}$ and $e_\alpha(\cdot)$ is a continuous
function on $\sphere^{d-1}$.

The second of the (id) conditions, which is satisfied when the
distribution of $A$ has a Lebesgue density, can actually be lifted if
stronger moment conditions are imposed on $A$ and $B$, and in
addition, a proximity condition is satisfied by the support of
$A$. This is the result of Guivarc'h and Le Page
\cite{guivarc:page:2016}. Let $G_A$ denote the semi-group generated
by $\{\Pi_n: \Pi_n = A_n \cdots A_1, A_i \in M(d, \reals)\}$. The
authors assumed
\begin{enumerate}
  \item There is no finite union $W$ of proper subspaces of $\reals^d$
    that satisfies $\forall a \in G_A, a W = W$.
  \item $G_A$ contains a proximal element, i.e. an element $a$ whose
    largest singular value is an algebraically simple eigenvalue of $a$.
\end{enumerate}
These two assumptions are termed (ip) conditions. Replacing the (id)
conditions of Alsmeyer and Mentemeier with (ip) and the moment
conditions of the former with
\[
\E (\|A\|^{\alpha + \delta}) < \infty, \quad
\E (\|A\|^{\alpha} \|A^{-1}\|^{\delta}) < \infty, \quad
\E (|B|^{\alpha + \delta}) < \infty \quad
\text{ for some }\delta > 0,
\]
Guivarc'h and Le Page proved the vague convergence result of
\eqref{eq:jorg}.

@@

\section{Stochastic volatility models}
\label{sec:SV:intro}
With the availability of high-frequency data, a different approach
than that of GARCH has been popularized for modelling volatility of
financial time series and has led to greatly improved accuracy of
prediction. This is the approach of stochastic volatility models.

In the pioneering work of Clark \cite{clark:1973}, the author modelled
the logarithmic price $Y_t$ as a subordinated stochastic process:
$Y_t = V_{\tau_t}$, where $V_i$ is a Brownian motion.
$\tau_t$ is a real-valued, non-negative, non-decreasing
sequence with $\tau_0 = 0$. It models a time change. As pointed out by
Shephard and Andersen \cite{Shephard:Andersen:2009}, the log-price
process $\{Y_t\}$ is serially uncorrelated although potentially dependent,
provided that $V_t$ and $\tau_t$ are independent.

Later authors, e.g. Back \cite{back:1991} chose to model the log-price
process as a semimartingale, with increments of the martingale
component modelled as a product process:
\begin{eqnarray}
  Y_t &=& Y_0 + A_t + M_t \nonumber \\
  M_t - M_{t-1} &=& X_t = \sigma_t Z_t
  \label{eq:rtght}
\end{eqnarray}
where $\{A_t\}$ is a finite-variation process and $\{M_t\}$ is a
martingale and hence $\{Y_t\}$ is a semimartingale.
$\{\sigma_t\}$ is non-negative and $Z_t$ is an iid process with zero mean
and unit variance. A convenient choice of $\sigma_t$ is
\begin{equation}
  \label{eq:rfht}
  \log\sigma_t = \sum_{l \in \integers} \psi_l \eta_{t-l}
  \quad
  t \in \integers,
\end{equation}
where $\{\psi_l\}_{l \in \integers}$ is a sequence of real numbers
with at least one non-zero element,
$\{\eta_t\}_{t \in \integers}$ is an iid sequence of random
variables. By Kolmogorov's 3-series theorem, the infinite series above
converges if and only if $\sum_{l \in \integers} \psi_l^2 < \infty$,
and in this case $\{\sigma_{t}\}$ is stationary.
In particular, if $\{\eta_i\}_{i \in \integers}$ is normally
distributed with zero mean, $\{\sigma_t\}_{t \in \integers}$ has
log-normal marginal distributions.

If, however, the sequences $\{Z_t\}_{t \in \integers}$ or
$\{\sigma_t\}_{t \in \integers}$ are regularly varying with index
$\alpha$ and some additional conditions are satisfied, $\{X_t\}$
is also regularly varying with the same index. Specifically,
if $\{Z_t\}$ is regularly varying and $\{\sigma_t\}$ has a lighter
tail, the conclusion follows from Breiman's lemma \cite{breiman:1965}.
See \S\ref{sec:case1} of Jan\ss en et al. i.e. chapter
\ref{ch:bernoulli} of this thesis for more details.

@@

\subsection{Extreme value analysis for the sample autocovariance matrices of time series}
Jan\ss en et al. \cite{janssen:mikosch:rezapour:xie:2016} investigated
the sample covariance matrix
$\mathbf X = \{\sum_{t=1}^n X_{i,t} X_{j,t}\}_{1 \leq i,j \leq p}$ of
multivariate {\em stochastic volatility models} assuming the dimension
of the matrix $p$ is fixed. It is also of interest to look into the
sample covariance and the sample auto-covariance matrix when the dimension
$p$ tends to infinity at some rate as the number of observations $n$ tends
to infinity. This is the subject of Davis et al.
\cite{davis:mikosch:heiny:xie:2015} which we summarize in this section.

>>>>

\chapter{Introduction}\label{ch:intr}
The phenomenon of heavy-tailedness is widely observed in many
disciplines of science, for example, phase transition of matter and
black body radiation as studied in physics, neuronal avalanches in
biology, claim sizes of insurance mathematics and stock returns in
finance. The last application is indeed the focus of this thesis. To
discuss the phenomena in precise terms, we introduce the concept of
regular variation.

\section{Regular Variation}\label{sec:intro_regvar}
The concept of regular variation is defined by the following scaling
property: if a function $f$ on $(0, \infty)$ satisfies
\[
\lim_{x \to \infty} {f(c x) \over f(x)} = c^\alpha
\quad
\forall c > 0,
\]
then we say $f$ is regularly varying with index $\alpha$.
$f$ can be written in the form $f(x) = {L(x) \over x^\alpha}$, where
$L(x)$ is a slowly varying function, i.e.
$\lim_{x \to \infty} {L(c x) \over L(x)} = 1$, $\forall c > 0$.
We call a random variable $X$ regularly varying with index
$\alpha \geq 0$ if it satisfies the tail balance condition in the
limit $x \to \infty$.
\[
\P(X > x) \sim p_+ {L(x) \over x^\alpha},\quad
\P(X < -x) \sim p_- {L(x) \over x^\alpha}, \quad
\text{for some } p_+, p_- \geq 0,\; p_+ + p_- = 1
\]

@@

Clearly, estimating the tail index $\alpha$ of a sequence
$X_1, X_2, \dots$ of regularly varying variables is particularly
important for understanding the behavior of a heavy-tailed series. A
standard method proposed for this purpose is due to Hill
\cite{hill1975simple}:
\begin{equation}
  \label{eq:fhhty}
  \hat \alpha_H = \left[
    {1 \over k} \sum_{i=1}^k \log \left(
      X_{(i)} \over X_{(k+1)}
    \right)
  \right]^{-1},
\end{equation}
where $X_1, X_2, \dots, X_n$ is a sample whose tail index is the
subject of estimation, and $X_{(i)}$ is the $i$-th upper order
statistic of the sample. Several authors have contributed to showing
the weak consistency and asymptotic normality of the estimator
$\hat \alpha_H$, under the assumptions $k \to \infty, k/n \to 0$
as $n \to \infty$. See Theorem 6.4.6 of Embrechts et. al.
\cite{embrechts:klueppelberg:mikosch:1997} for details.

@@

Random variables with regularly varying tails have some very nice
features: if $X_1$ and $X_2$ are both positive and regularly varying
with indices $\alpha_1$ and $\alpha_2$, respectively, then $a_1 X_1 + a_2 X_2$,
for $a_1, a_2 > 0$, is regularly varying with index
$\min\{\alpha_1, \alpha_2\}$ (cf. Mikosch and Jessen
\cite{JessenMikosch2006}). Moreover, if $X_1, X_2$ are iid,
$\P(a_1 X_1 + a_2 X_2 > u) \sim \P(a_1 X_1 > u) + \P(a_2 X_2 > u)$.

Now consider $p$ return series
$X_{i,t}$, $i=1,2,\dots, p; t=1,\dots,n$.
Suppose each of these series is a linear combination of $K$ factors
$Y_{j,t}$, $j=1,2,\dots,K$, the $j$-th of which is regularly varying
with index $\alpha_j$. Then by the summation property, each and every
$\{X_{i,t}\}$ is regularly varying with index $\min_{1 \leq j \leq K} \alpha_j$.
In practice a factor $Y_{i,t}$ is estimated as
$\hat Y_{j,t} = \sum_{i=1}^p a_{j, i} X_{i,t}$, where
$(a_{j, 1}, a_{j, 2}, \dots, a_{j, p})'$ is the $j$-th eigenvector
of the sample covariance matrix of
$\{X_{i,t}\}, i=1,\dots,p; t=1,\dots,n$, i.e. the eigenvector
associated with the $j$-th largest eigenvalue. 
For this reason, it is important to understand the eigen system of the
sample covariance matrix of $\{X_{i,t}\}$. This topic is addressed in
chapters \ref{ch:bernoulli} and \ref{ch:extremes} of this thesis.

@@

The notion of a regularly varying strictly stationary process is also of
interest. Originally introduced by Davis and Hsing
\cite{davis:hsing:1995} for univariate processes, the concept was
extended to multivariate processes by Davis and Mikosch
\cite{davis:mikosch:2009b}: an $\reals^d$-valued strictly stationary
process $X_t$ is said to be regularly varying with index $\alpha$, if
for each $h \geq 0$, 
\[
  {
    \P\left[x^{-1}(X_0, X_1, \dots, X_h) \in \cdot \right]
    \over
    \P(|X_0| > x)
  } \overset{v}{\to} \nu^{(h)}_\alpha(\cdot),
  \quad x \to \infty
\]
where $\nu_\alpha^{(h)}$ a non-null Radon measure on
$\reals^{d(h+1)} \setminus \{0\}$ that
is homogeneous of order $-\alpha$, i.e.
$\forall c > 0, \forall S \subset \reals^{d(h+1)}\setminus\{0\}, \nu_\alpha^{(h)}(c S) = c^{-\alpha} \nu_\alpha^{(h)}(S)$. $|\cdot|$ could be any given vector norm.

Basrak and Segers \cite{basrak:segers:2009} proved an equivalence
condition for regular variation of
a strictly stationary process: an $\reals^d$-valued strictly stationary
process $\{X_t\}$ is regularly varying with index $\alpha$ if and only if
there exists an $\reals^d$-valued sequence $\{\Theta_i\}_{i = 0,1,\ldots}$
and a Pareto($\alpha$) random variable $Z$, i.e.
$\P(Z > x) = x^{-\alpha}, \forall x \geq 1$, independent of
$\{\Theta_i\}$
\[
\P\left[\left.
  x^{-1}(X_0, X_1, \ldots, X_h) \in \cdot
  \right| |X_0| > x
  \right]
  \overset{w}{\to}
  \P(Z(\Theta_1, \ldots, \Theta_h) \in \cdot),
  \quad
  x \to \infty, \forall h \geq 0
\]
The sequences $\{\Theta_i\}_{i = 0,1,\ldots}$ and
$\{Z \Theta_i\}_{i = 0,1,\ldots}$ are called the spectral tail process
and the tail process of $\{X_t\}$, respectively.

A recent remarkable result from Mikosch and Wintenberger
\cite{mikosch:wintenberger:2016} regarding
strictly stationary regularly varying processes is quoted below:
\begin{theorem}\label{thm:mikwin:intro}
Let $(Y_t)$ be an $\bbr^d$-valued strictly stationary sequence , $S_n=Y_1+\cdots +Y_n$ and $(a_n)$ be such that $n\,\P(|Y|>a_n)\to 1$.
Also write for $\vep>0$, $\ov Y_t = Y_t \I(|Y_t|\le \vep a_n)$, $\underline Y_t= Y_t-\ov  Y_t $ and
\beao
\ov S_{l,n}= \sum_{t=1}^l \ov Y_t \qquad  \un S_{l,n}= \sum_{t=1}^l \un Y_t\,.
\eeao
Assume the following conditions:
\begin{enumerate}
\item 
$(Y_t)$ is regularly varying with index $\alpha\in (0,2)  \setminus\{1\}$ and spectral tail process $(\Theta_j)$.
\item
A mixing condition holds: there exists an integer \seq\ $m_n\to\infty$ \st\ $k_n= [n/m_n]\to \infty$
and 
\beam\label{eq:chfa:intro}
\E \ex^{it'\underline S_n/a_n} - \Big(\E \ex^{it'\underline S_{m_n,n}/a_n}\Big)^{k_n} \to 0\,,\qquad \nto\,,\qquad t\in\bbr^d\,.
\eeam
\item An anti-clustering condition holds:
\beam\label{eq:acl:intro}
\lim_{l\to\infty} \limsup_{\nto} \P\big(\max_{t=l,\ldots,m_n} |Y_t|>\delta a_n\mid |Y_0|>\delta a_n\big)=0\,,\qquad \delta>0\,
\eeam
for the same sequence $(m_n)$ as in 2.
\item If $\alpha\in (1,2)$, in addition $\E[Y]=0$ and the vanishing small values condition holds: 
\beam\label{eq:vansm:intro}
\lim_{\vep\downarrow 0}\limsup_{\nto} \P\big(a_n^{-1} |\ov S_n-\E [\ov S_n]|>\delta \big)=0\,,\qquad \delta>0\,
\eeam
and $\sum_{i=1}^\infty \E[|\Theta_i|]<\infty$.
\end{enumerate}
Then $a_n^{-1} S_n\std \xi_\alpha$ for an $\alpha$-stable $\bbr^d$-valued
vector $\xi_\alpha$ with log characteristic function
\begin{small}
  \beam\label{eq:chfid:intro}
  \int_0^\infty \E\left[
    \exp\left(i\,y\,t'\sum_{j=0}^\infty \Theta_j\right)-
    \exp\left(i\,y\,t'\sum_{j=1}^\infty \Theta_j\right)
    -i\,y\,t'\I_{(1,2)}(\alpha)
    \right]\,
  d(-y^{\alpha})\,,\qquad t\in\bbr^d\,.
  \eeam
\end{small}
\end{theorem}\noindent 
\begin{remark}\label{rem:case:alpha:1:intro}
If we additionally assume that $Y$ is symmetric, which implies $\E[\ov
  Y]=\mathbf{0}$, then the statement of the theorem also holds for
$\alpha=1$.
\end{remark}

@@

Kesten \cite{kesten:1973} showed that, when $A_t$ is almost
surely non-negative, has no row or column of only zeros, and
$B_t$ is almost surely non-negative and is not equal to the null
vector with probability 1, then the solution $X$ to the equation
$X \overset{d}{=} A X + B$ is regularly varying with a positive index
$\alpha$, assuming the following conditions (M) and (A):
\begin{itemize}
\item Condition (M)
  \begin{enumerate}
  \item The top Lyapunov exponent
    \[
    \gamma = \inf_{n \geq 1} {1 \over n}\E \log \|A_n \cdots A_1\|
    \]
    is negative.
  \item There exists $\alpha > 0$ such that
    \[
    1 = \lambda(\alpha) = \lim_{n \to \infty} {1 \over n} \log \E
    \|A_n \cdots A_1\|^\alpha
    \]
  \item $\E (\|A_1\|^\alpha \log^+\|A_1\|) < \infty$
  \item $\E |B_1|^\alpha < \infty$
  \end{enumerate}
\item Condition (A) : the group generated by
  \[
  \{\log\rho(s): s = A_n \cdots A_1 \text{ for some } n \geq 1\}
  \]
  is dense in $\reals$, where $\rho(s)$ denotes the spectral
  radius of matrix $s$.
\end{itemize}
Upon these conditions, Kesten's theorem gives
\begin{equation}
  \label{eq:jorg}
  u^\alpha \P(u^{-1} X \in \cdot) \overset{v}{\to} \nu_\alpha^{(h)}(\cdot)
\end{equation}
where $\nu_\alpha^{(h)}$ is a non-null Radon measure on
$\reals^d_+ \setminus \{0\}$ with the property
$\forall a > 0, \nu_\alpha^{(h)}(a A) = a^{-\alpha} \nu_\alpha^{(h)}(A)$.

In addition to non-negative matrices, two other classes of random
matrices have been shown to lead to power-law tails via the recurrence
relation \eqref{eq:garchpq_sre}. Alsmeyer and Mentemeier
\cite{alsmeyer:mentemeier:2012} considered invertible matrices whose
distribution has a density. Let $M(d, \reals)$ denote a metric space of
$d \times d$ matrices with real entries that are invertible with
probability 1. They replaced Kesten's condition
$\E (\|A\|^\alpha \log^+\|A\|) < \infty$ with a stronger
counterpart
$\E [\|A\|^\alpha (\log^+\|A\| + \log \|A^{-1} \|)] < \infty$,
and lifted the condition (A). In addition, they assumed
\begin{enumerate}
  \item For any open set $U \subset \sphere^{d-1}$ and any
    $u \in \sphere^{d-1}$,  $\exists n \geq 1$ such that
    \[
    \P\left({\prod_{i=1} A_i u
        \over
        \left|\prod_{i=1} A_i u\right|}\in U
    \right) > 0.
    \]
  \item There exist $N \geq 1$, $c, \epsilon > 0$ and an invertible
    matrix $\bar A \in M(d, \reals)$ such that for any set
    $C \subset M(d, \reals)$, it holds true
    $\P(A_N \cdots A_1 \in C) \geq c |B_\epsilon(\bar A) \cap C|$,
    where $|\cdot|$ denotes the Lebesgue measure and
    $B_\epsilon(\bar A)$ is the ball with radius $\epsilon$ centered
    at $\bar A$.
\end{enumerate}
These assumptions are termed conditions (id). Furthermore, they assumed
that there was no point in $\reals^d$ such that the recurrence
equation \eqref{eq:garchpq_sre} was stuck at this point with probability 1: 
$\P(A x + B = x) < 1$ for all $x \in \reals^d$. With these
assumptions, they showed
\[
\lim_{u \to \infty} u^\alpha \P(\inn{x, X} > u) = e_\alpha(x)
\]
where $x \in \sphere^{d - 1}$ and $e_\alpha(\cdot)$ is a continuous
function on $\sphere^{d-1}$.

The second of the (id) conditions, which is satisfied when the
distribution of $A$ has a Lebesgue density, can actually be lifted if
stronger moment conditions are imposed on $A$ and $B$, and in
addition, a proximity condition is satisfied by the support of
$A$. This is the result of Guivarc'h and Le Page
\cite{guivarc:page:2016}. Let $G_A$ denote the semi-group generated
by $\{\Pi_n: \Pi_n = A_n \cdots A_1, A_i \in M(d, \reals)\}$. The
authors assumed
\begin{enumerate}
  \item There is no finite union $W$ of proper sub-spaces of $\reals^d$
    that satisfies $\forall a \in G_A, a W = W$.
  \item $G_A$ contains a proximal element, i.e. an element $a$ whose
    largest singular value is an algebraically simple eigenvalue of $a$.
\end{enumerate}
These two assumptions are termed (ip) conditions. Replacing the (id)
conditions of Alsmeyer and Mentemeier with (ip) and the moment
conditions of the former with
\[
\E (\|A\|^{\alpha + \delta}) < \infty, \quad
\E (\|A\|^{\alpha} \|A^{-1}\|^{\delta}) < \infty, \quad
\E (|B|^{\alpha + \delta}) < \infty \quad
\text{ for some }\delta > 0,
\]
Guivarc'h and Le Page proved the vague convergence result of
\eqref{eq:jorg}.

@@

\section{Stochastic volatility models}
\label{sec:SV:intro}
With the availability of high-frequency data, a different approach
than that of GARCH has been popularized for modeling volatility of
financial time series and has led to greatly improved accuracy of
prediction. This is the approach of stochastic volatility models.

In the pioneering work of Clark \cite{clark:1973}, the author modeled
the logarithmic price $Y_t$ as a subordinated stochastic process:
$Y_t = V_{\tau_t}, t\geq 0$, where $V_i$ is a Brownian motion.
$\tau_t$ is a real-valued, non-negative, non-decreasing
sequence with $\tau_0 = 0$. It models a time change. As pointed out by
Shephard and Andersen \cite{Shephard:Andersen:2009}, the log-price
process $\{Y_t\}$ is serially uncorrelated although potentially dependent,
provided that $V_t$ and $\tau_t$ are independent.

Later authors, e.g. Back \cite{back:1991} chose to model the log-price
process as a semi-martingale, with increments of the martingale
component modeled as a product process:
\begin{eqnarray*}
  Y_t &=& Y_0 + A_t + M_t,
\end{eqnarray*}
where $\{A_t\}$ is a finite-variation process and $\{M_t\}$ is a
martingale and hence $\{Y_t\}$ is a semi-martingale.
$\{\sigma_t\}$ is non-negative and $\{Z_t\}$ is an iid process with
zero mean and unit variance. $\{\sigma_t\}$ and $\{Z_t\}$ are
independent of each other.

For discrete time models, Taylor \cite{taylor:1982} was the first to
propose a product process for modeling the martingale part $M_t$ of
$Y_t$:
\[
M_t - M_{t-1}= \sigma_t Z_t, \quad t \in \integers
\]
A convenient choice of $\sigma_t$ is
\begin{equation}
  \label{eq:rfht}
  \log\sigma_t = \sum_{l \in \integers} \psi_l \eta_{t-l}
  \quad
  t \in \integers,
\end{equation}
where $\{\psi_l\}_{l \in \integers}$ is a sequence of real numbers
with at least one non-zero element,
$\{\eta_t\}_{t \in \integers}$ is an iid sequence of random
variables with zero-mean and finite variance.
By Kolmogorov's 3-series theorem, the infinite series above
converges if and only if $\sum_{l \in \integers} \psi_l^2 < \infty$,
and in this case $\{\sigma_{t}\}$ is stationary.
In particular, if $\{\eta_i\}_{i \in \integers}$ is normally
distributed with zero mean, $\{\sigma_t\}_{t \in \integers}$ has
log-normal marginal distributions.

If, however, the sequences $\{Z_t\}_{t \in \integers}$ or
$\{\sigma_t\}_{t \in \integers}$ are regularly varying with index
$\alpha$ and some additional conditions are satisfied, $\{X_t\}$
is also regularly varying with the same index. Specifically,
if $\{Z_t\}$ is regularly varying and $\{\sigma_t\}$ has a lighter
tail, the conclusion follows from Breiman's lemma \cite{breiman:1965}.
See \S\ref{sec:case1} of Jan\ss en et
al. \cite{janssen:mikosch:rezapour:xie:2016}, i.e. chapter
\ref{ch:bernoulli} of this thesis for more details.

@@

\subsection{Extreme value analysis for the sample auto-covariance
  matrices of time series}
Jan\ss en et al. \cite{janssen:mikosch:rezapour:xie:2016} investigated
the sample covariance matrix of the time series
$\{X_{i,t}\}_{t \in \integers}$, $1 \leq i \leq p$ for
{\em stochastic volatility models} assuming the dimension
of the matrix $p$ is fixed. It is also of interest to look into the
sample covariance and the sample auto-covariance matrix when the dimension
$p$ tends to infinity at some rate as the number of observations $n$ tends
to infinity. This is the subject of Davis et al.
\cite{davis:mikosch:heiny:xie:2015} which we summarize in this section.
